---
output: 
  html_document:
    toc: true
    code_folding: "hide"
---

# Getting started

```{r, echo=FALSE}
suppressPackageStartupMessages(library(SummarizedExperiment))
if(!require(airway)) {
    BiocInstaller::biocLite('airway')
}
suppressPackageStartupMessages(data(airway, package="airway"))
```

```{r}
library(SummarizedExperiment)
data(airway, package="airway")
airway
```

# Exercises

- What is the `class()` of `airway`? What does the class tell you about the information 
stored about the features (genes) in the dataset?

```{r results="hide"}
class(airway)
```

- How many features are in the data?

```{r results="hide"}
airway # it tells you
# but to get the result as a number:
nrow(airway)
```

- How many samples are present?

```{r results="hide"}
airway # it tells you
# but to get the result as a number:
ncol(airway)
```

- The assay information in the airway object contains the RNA-seq counts. Access the count data for the 
56th gene. 

```{r results="hide"}
# check the names of the available assays
assayNames(airway)
# Both below are identical
assay(airway, "counts")[56, ]
assays(airway)[[1]][56, ]
```

- What are the `rownames()` of airway? What do these represent? What gene is represented by the first row
of data? 

```{r results="hide"}
head(rownames(airway))
# These are Ensembl gene identifiers
# ENSG00000000003 is the identifier of the "TSPAN6" gene
```

- What information is available about the samples? 

```{r results="hide"}
colData(airway)
```

- How many samples were treated with dex? Untreated? 

```{r results="hide"}
colData(airway) # and count
colData(airway)$dex # and count
table(colData(airway)$dex)
```

- How many different cell types were used?

```{r results="hide"}
table(colData(airway)$cell)
length(unique(colData(airway)$cell))
```

- What does `rowRanges(airway)` give you? What is the length of the object? What does each element of
`rowRanges(airway)` contain? 

```{r results='hide'}
rowRanges(airway)
length(rowRanges(airway))
rowRanges(airway)[[1]]
```

- From the [PubMed abstract](https://www.ncbi.nlm.nih.gov/pubmed/24926665), we can see that the authors
suggest that the DUSP1 gene is upregulated by Dexamethasone treatment. The [GeneCards website](https://www.genecards.org/cgi-bin/carddisp.pl?gene=DUSP1&keywords=DUSP1) tells us that the 
Ensembl gene id for DUSP1 is 'ENSG00000120129'. Get the count data for this gene.

```{r results="hide"}
# get the assay data for DUSP1
dat = assay(airway, "counts")["ENSG00000120129", ]
```

- Make a plot of the gene counts in the previous exercise. Color the points based on the dexamethasone treatment.

```{r results="hide"}
# get the assay data for DUSP1
plot(dat)
dextrt = colData(airway)$dex
plot(dat, col=dextrt)
```

# sessionInfo()

```{r}
sessionInfo()
```