---
title: "Data Frames"
output: html_document
---

While R has many different data types, the one that is central to much of the power
and popularity of R is the `data.frame`. A data.frame looks a bit like an R matrix in
that it has two dimensions, rows and columns. However, data.frames are usually viewed
as a set of columns representing variables and the rows representing the values of
those variables. Importantly, a data.frame may contain *different* data types in each 
of its columns; matrices **must** contain only one data type. This distinction is 
important to remember, as there are *specific* approaches to working with R data.frames
that may be different than those for working with matrices.

## Dataset

The data used here are borrowed directly from the [fantastic Bioconnector tutorials](http://bioconnector.org/index.html) and are a cleaned up
version of the data from [Brauer et al. Coordination of Growth Rate, Cell Cycle, Stress Response, and Metabolic Activity in Yeast (2008) Mol Biol Cell 19:352-367](http://www.ncbi.nlm.nih.gov/pubmed/17959824). These data are from a gene expression microarray, and in this paper the authors examine the relationship between growth rate and gene expression in yeast cultures limited by one of six different nutrients (glucose, leucine, ammonium, sulfate, phosphate, uracil). If you give yeast a rich media loaded with nutrients except restrict the supply of a single nutrient, you can control the growth rate to any rate you choose. By starving yeast of specific nutrients you can find genes that:

1. Raise or lower their expression in response to growth rate. Growth-rate dependent expression patterns can tell us a lot about cell cycle control, and how the cell responds to stress. The authors found that expression of >25% of all yeast genes is linearly correlated with growth rate, independent of the limiting nutrient. They also found that the subset of negatively growth-correlated genes is enriched for peroxisomal functions, and positively correlated genes mainly encode ribosomal functions.
2. Respond differently when different nutrients are being limited. If you see particular genes that respond very differently when a nutrient is sharply restricted, these genes might be involved in the transport or metabolism of that specific nutrient.

The dataset can be downloaded directly from:

- [http://bioconnector.org/data/brauer2007_tidy.csv](http://bioconnector.org/data/brauer2007_tidy.csv)

We are going to read this dataset into R and then use it as a playground for learning about data.frames.

## Reading in data

R has many capabilities for reading in data. Many of the functions have names that help us 
to understand what data format is to be expected. In this case, the filename that we want to
read ends in `.csv`, meaning comma-separated-values. The `read.csv()` function reads in `.csv` files.
As usual, it is worth reading `help('read.csv')` to get a better sense of the possible
bells-and-whistles. 

The `read.csv()` function can read directly from a URL, so we do not need to download the file directly. 
This dataset is relatively large (about 16MB), so this may take a bit depending on your network connection 
speed.

```{r cache=TRUE}
ydat <- read.csv('http://bioconnector.org/data/brauer2007_tidy.csv')
```

Our variable, `ydat`, now "contains" the downloaded and read data. We can check to see what data type 
read.csv gave us:

```{r}
class(ydat)
```

## Inspecting data.frames

Our `ydat` variable is a data.frame. As I mentioned, the dataset is fairly large, so we will not be
able to look at it all at once on the screen. However, R gives us many tools to inspect a data.frame.

- Overviews of content
  + `head()` to show first few rows
  + `tail()` to show last few rows
- Size
  + `dim()` for dimensions (rows, columns)
  + `nrow()`
  + `ncol()`
  + `object.size()` for power users interested in the memory used to store an object
- Data and attribute summaries
  + `colnames()` to get the names of the columns
  + `rownames()` to get the "names" of the rows--may not be present
  + `summary()` to get per-column summaries of the data in the data.frame.
  
```{r dataframeinspection}
head(ydat)
tail(ydat)
dim(ydat)
nrow(ydat)
ncol(ydat)
colnames(ydat)
summary(ydat)
```

In RStudio, there is an additional function, `View()` (note the capital "V") that opens the first
1000 rows (default) in the RStudio window, akin to a spreadsheet view. 

```{r eval=FALSE}
View(ydat)
```

## Accessing variables (columns) and subsetting

In R, data.frames can be subset similarly to other two-dimensional data structures. The `[` in R is used 
to denote subsetting of any kind. When working with two-dimensional data, we need two values inside the `[ ]`
to specify the details. The specification is `[rows, columns]`. For example, to get the first three rows of `ydat`,
use:

```{r}
ydat[1:3, ]
```

Note how the second number, the columns, is blank. R takes that to mean "all the columns". Similarly, we can
combine rows and columns specification arbitrarily.

```{r}
ydat[1:3, 1:3]
```

Because selecting a single variable, or column, is such a common operation, there are two shortcuts for
doing so *with data.frames*. The first, the `$` operator works like so:

```{r}
# Look at the column names, just to refresh memory
colnames(ydat)
# Note that I am using "head" here to limit the output
head(ydat$symbol)
# What is the actual length of "symbol"?
length(ydat$symbol)
```

The second is related to the fact that, in R, data.frames are also lists. We subset a list by using `[[]]` notation.
To get the second column of `ydat`, we can use:

```{r}
head(ydat[[2]])
```

Alternatively, we can use the column name:

```{r}
head(ydat[["systematic_name"]])
```

### More advanced indexing and subsetting

```{r}
head(ydat[ydat$symbol == 'LEU1', ])
dim(ydat[ydat$expression > 3, ])
```

Find the high expressed genes when leucine-starved. For this task we can also use `subset` which allows 
us to treat column names as R variables (no `$` needed).

```{r}
subset(ydat, nutrient == 'Leucine' & rate == 0.05 & expression > 3)
```

## Aggregating data

```{r}
aggregate(ydat$expression, by=list( ydat$symbol), mean)
# or 
aggregate(expression ~ symbol, mean, data=ydat)
```

## Creating a data.frame from scratch

Sometimes it is useful to combine related data into one object. For example, let's simulate some data.

```{r}
smoker = factor(rep(c("smoker", "non-smoker"), each=50))
smoker_numeric = as.numeric(smoker)
x = rnorm(100)
risk = x + 2*smoker_numeric
```

We have two varibles, `risk` and `smoker` that are related. We can make a data.frame out of them:

```{r}
smoker_risk = data.frame(smoker = smoker, risk = risk)
head(smoker_risk)
```

R also has plotting shortcuts that work with data.frames to simplify plotting

```{r}
plot( risk ~ smoker, data=smoker_risk)
```

## Saving a data.frame

Once we have a data.frame of interest, we may want to save it. The most
portable way to save a data.frame is to use one of the `write` functions. In
this case, let's save the data as a `.csv` file.

```{r eval=FALSE}
write.csv(smoker_risk, "smoker_risk.csv")
```


