---
title: "Data Frames"
output: html_document
---

```{r}
library(knitr)
opts_chunk$set(eval=FALSE)
```

## Matrices Versus Data Frames


```{r}
mat<-cbind(x,y)
head(mat)
class(mat[,1])          
z = paste0('a',1:10)
tab<-cbind(x,y,z)
class(tab)
mode(tab[,1])
head(tab,4)
```


```{r}
tab<-data.frame(x,y,z)
class(tab)
head(tab)
mode(tab[,1])
class(tab[,3])
rownames(tab)           
rownames(tab)<-paste0("row",1:10)
rownames(tab)
```




Data frame columns can be refered to by name using the "dollar sign" operator.

```{r}
head(tab)
tab$x
tab$y
```

Column names can be set, which can be useful for referring to data
later.

```{r}
colnames(tab)
colnames(tab) = paste0('col',1:3)
```

Data frames have functions to report size and even `summary` functions. Try the following:

```{r}
ncol(tab)
nrow(tab)
dim(tab)
summary(tab)
```

Extracting parts of a `data.frame` work as for matrices. Try to think about what each of the following will do before asking R to evaluate the result. 
```{r}
tab[1:3,]
tab[,2:3]
tab[,1]>7
tab[tab[,1]>7,]
tab[tab[,1]>7,3]
tab[tab[,1]>7,2:3]
tab[tab$x>7,3]
tab$z[tab$x>3]
```

## Lists and Objects

A list is a collection of objects that may be the same or different
types. The objects in a list often have names, and may be indexed either by name
(e.g. `my_list$name3`) or component number (e.g. `my_list[[3]]`). 
As an aside, a data frame is a list of matched column vectors.

We can reate a list, noting the different data types involved.

```{r}
a = list(1,"b",c(1,2,3))
a
length(a)
class(a)
a[[3]]
```

A data frame *is* a list.

```{r}
# test if our friend "tab" is a list
is.list(tab)
tab[[2]]
names(tab)
```

Table (\#tab:datatypes) Summary of basic R data structures.

  Data type   Stores
  ------------ ----------------------------------------------------------
     vector              one-dimensional data, single data type
     matrix              two-dimensional data, single data type
   data frame          two-dimensional data, multiple data types
      list        list of data types, not all need to be the same type
     object     a list with attributes and potentially slots and methods

