---
title: "Data exploration and univariate statistics"
---

<!-- Thanks to Martin Morgan for much of this material! -->

```{r style-A3, echo = FALSE, results = 'asis'}
knitr::opts_chunk$set(
    eval=as.logical(Sys.getenv("KNITR_EVAL", "TRUE")),
    cache=as.logical(Sys.getenv("KNITR_CACHE", "TRUE")))
```

## Behavioral Risk Factor Surveillance System

We will explore a subset of data collected by the CDC through its
extensive Behavioral Risk Factor Surveillance System ([BRFSS][])
telephone survey. Check out the link for more information. We'll look
at a subset of the data.

First, we need to get the data. Either download the data
from [THIS LINK](BRFSS-subset.csv) or have R do it directly from the
command-line (preferred):

```{r}
download.file('https://raw.githubusercontent.com/seandavi/ITR/master/BRFSS-subset.csv',
              destfile = 'BRFSS-subset.csv')

```


<!--
```{r echo=FALSE}
path <- "BRFSS-subset.csv"
```
-->
```{r ALL-choose-A3, eval=FALSE}
path <- file.choose()    # look for BRFSS-subset.csv
```

```{r ALL-input-A3}
stopifnot(file.exists(path))
brfss <- read.csv(path)
```

## Clean data

_R_ read `Year` as an integer value, but it's really a `factor`

```{r}
brfss$Year <- factor(brfss$Year)
```

## Weight in 1990 vs. 2010 Females

- Create a subset of the data

```{r}
brfssFemale <- brfss[brfss$Sex == "Female",]
summary(brfssFemale)
```

- Visualize

```{r}
plot(Weight ~ Year, brfssFemale)
```

- Statistical test

```{r}
t.test(Weight ~ Year, brfssFemale)
```

## Weight and height in 2010 Males

- Create a subset of the data

```{r}
brfss2010Male <- subset(brfss,  Year == 2010 & Sex == "Male")
summary(brfss2010Male)
```

- Visualize the relationship

```{r}
hist(brfss2010Male$Weight)
hist(brfss2010Male$Height)
plot(Weight ~ Height, brfss2010Male)
```

- Fit a linear model (regression)

```{r}
fit <- lm(Weight ~ Height, brfss2010Male)
fit
```

Summarize as ANOVA table

```{r}
anova(fit)
```

- Plot points, superpose fitted regression line; where am I?

```{r}
plot(Weight ~ Height, brfss2010Male)
abline(fit, col="blue", lwd=2)
# Substitute your own weight and height...
points(73 * 2.54, 178 / 2.2, col="red", cex=4, pch=20)
```

- Class and available 'methods'

```{r, eval=FALSE}
class(fit)                 # 'noun'
methods(class=class(fit))  # 'verb'
```

- Diagnostics

```{r, eval=FALSE}
plot(fit)
?plot.lm
```

