---
title: "Swirl"
output: html_document
---

The following is from the [swirl website](http://swirlstats.com/students.html).

> The swirl R package makes it fun and easy to learn R programming and data science. If you are new to R, have no fear. 

Swirl is an interactive system for learning *R*, written and used in *R*. 

## Installation

To get started, we need to install a new package into R. 

```{r eval=FALSE}
install.packages('swirl')
```

## Use

Once installed, we want to load it into the R workspace so we can use it.

```{r eval=FALSE}
library('swirl')
```

Finally, to get going, start swirl and follow the instructions.

```{r eval=FALSE}
swirl()
```

## Installing courses

The first time you start swirl, you'll be prompted to install a course. You can either install one of the recommended courses or visit the [course repository](https://github.com/swirldev/swirl_courses#swirl-courses) for more options. There are even more courses available from the Swirl Course Network.

If you'd like to install a course that is not part of our course repository, type `?InstallCourses` at the R prompt for a list of functions that will help you do so.

