---
title: "Vector Exercises"
output: 
  html_document:
    code_folding: "hide"
---

```{r include=FALSE}
library(knitr)
opts_chunk$set(results='hide')
```

## Exercises

- Create a numeric vector that starts at 1 and goes to 10 and assign it to the variable v10.

```{r}
v10 = 1:10
length(v10)
```

- Read the help for the `rnorm` function. What does this function do? Use the `rnorm` function to sample 100 values from
the normal distribution and assign it to the variable my_rn.

```{r}
my_rn = rnorm(100)
length(rnorm)
```

- What are the mean and standard deviation of `my_rn`? Consider using the `mean` and `sd` functions. Read the 
help for these functions as well.

```{r}
mean(my_rn)
sd(my_rn)
# another useful summary
summary(my_rn)
```

- How could you change the mean of `my_rn` from its current value to its current value + five?

```{r}
# a couple of options
# Add 5 to every value of my_rn
my_rn2 = my_rn + 5
mean(my_rn2)
```

- You are all going to present your work during the course. We want to do this as fairly as possible, so we want
to use R to randomly assign an order to you all. If each of you has a number between 1 and 25, how can you produce
a randomized ordering to see who goes first, second, etc.? See the `sample()` function for some help. Assign the ordering to a variable `speakers`.

```{r}
speakers = sample(1:25, replace = FALSE)
# what happens if you use replace=TRUE, the default?
```

- Create two new vectors, `day1` and `day2` containing the first 13 speakers and the last 12 speakers:

```{r}
day1 = speakers[1:13]
day2 = speakers[14:25]
```

- Use the `rep()` function to create a vector of length 10 with just "A" repeating. Do the same, but with "A" and "B" repeating. 

```{r}
# This will work
rep("A", 10)
# What is wrong with this one?
rep(c("A", "B"), 10)
# And how is this different?
# How would you know that you need to include
#   "each"? Read the help, of course. (or trial and error)
rep(c("A", "B"), each=5)
```


- Referring back to your `my_rn` example, use the `<` operator to create a vector of `TRUE` and `FALSE` 
so that you isolate values more than 1 standard deviation below the mean. 
Finally, use the vector of `TRUE`/`FALSE` to get the 
actual values from `my_rn` that meet these criteria. This is a multi-step problem, so 
think about the pieces you need and about how to put them together.

```{r}
# first determine the mean
mean(my_rn)
# then, find the sd
sd(my_rn)
# What is one sd below the mean?
mean(my_rn) - sd(my_rn)
# Assign to a variable
below_mean = mean(my_rn) - sd(my_rn)
# Do the comparison
my_rn < below_mean
# We can even capture to a variable
below_mean_index = my_rn < below_mean
# Finally, we can check our answer
my_rn[below_mean_index]
```

- Make a histogram of your `my_rn` vector (see `hist`). Add a vertical red line to the plot at 1 sd above and below the mean (see `help('abline')` and a green line at the mean.

```{r}
hist(my_rn)
abline(v = mean(my_rn) - sd(my_rn), col = 'red')
abline(v = mean(my_rn) + sd(my_rn), col = 'red')
abline(v = mean(my_rn), col = 'green')
```

## Further examples

If you would like further examples and interactive exercises, try the free [DataCamp vector exercises](https://campus.datacamp.com/courses/free-introduction-to-r/chapter-2-vectors-2).
