---
title: "Data Input and Manipulation Exercises"
output: 
  html_document:
    code_folding: "hide"
---

```{r style-A2, echo = FALSE, results = 'asis'}
knitr::opts_chunk$set(
    eval=as.logical(Sys.getenv("KNITR_EVAL", "TRUE")),
    cache=as.logical(Sys.getenv("KNITR_CACHE", "TRUE")))
```

## Behavioral Risk Factor Surveillance System

We will explore a subset of data collected by the CDC through its
extensive Behavioral Risk Factor Surveillance System ([BRFSS][])
telephone survey. Check out the link for more information. We'll look
at a subset of the data.

First, we need to get the data. Either download the data
from [THIS LINK](BRFSS-subset.csv) or have R do it directly from the
command-line (preferred):

```{r}
download.file('https://raw.githubusercontent.com/seandavi/ITR/master/BRFSS-subset.csv',
              destfile = 'BRFSS-subset.csv')

```

You can check to see the file using the Rstudio file panel or get a
directory listing using `dir()`

1. Use `file.choose()` to find the path to the file 'BRFSS-subset.csv'. This is a quick-and-dirty to 
find a file on the computer. Store the file location into a variable called `path`. 

    ```{r file.choose-A2, eval=FALSE}
    path <- file.choose()
    ```

<!--
    ```{r file.path, echo=FALSE}
    path <- "BRFSS-subset.csv"
    ```
-->

2. Read the data into R using `read.csv()`, assigning to a variable `brfss`. Note that you can use the `path` 
variable in read.csv().

    ```{r read.csv-A2}
    brfss <- read.csv(path)
    ```
    

3. Use command like `class()`, `head()`, `dim()`, `summary()` to explore
the data.

    - What variables have been measured?

    - Can you guess at the units used for, e.g., Weight and Height?
    
    ```
    class(brfss)
    head(brfss)
    dim(brfss)
    summary(brfss)
    ```

4. Use the `$` operator to extract the `Sex` column, and summarize the
   number of males and females in the survey using `table(brfss$Sex)`. Do the
   same for `Year`, and for both `Sex` and `Year`. 

    ```{r brfss-sex-A2}
    table(brfss$Sex)
    table(brfss$Year)
    table(brfss$Sex, brfss$Year)
    with(brfss, table(Sex, Year))                # same, but easier
    ```

5. Use `aggregate()` to summarize the mean weight of each group. What
   about the median weight of each group? What about the _number_ of
   observations in each group?

    ```{r brfss-aggregate}
    with(brfss, aggregate(Weight, list(Year, Sex), mean, na.rm=TRUE))
    with(brfss, aggregate(Weight, list(Year=Year, Sex=Sex), mean, na.rm=TRUE))
    ```

6. Use a `formula` and the `aggregate()` function to describe the
   relationship between Year, Sex, and Weight

    ```{r brfss-aggregate-formula}
    aggregate(Weight ~ Year + Sex, brfss, mean)  # same, but more informative
    aggregate(. ~ Year + Sex, brfss, mean)       # all variables
    ```

7. Create a subset of the data consisting of only the 1990
   observations. Perform a t-test comparing the weight of males and
   females ("'Weight' as a function of 'Sex'", `Weight ~ Sex`)

    ```{r t-test-1990}
    brfss_1990 = brfss[brfss$Year == 1990,]
    t.test(Weight ~ Sex, brfss_1990)
    t.test(Weight ~ Sex, brfss, subset = Year == 1990)
    ```
   What about differences between weights of males (or females) in
   1990 versus 2010? Check out the help page `?t.test.formula`. Is
   there a way of performing a t-test on `brfss` without explicitly
   creating the object `brfss_1990`?

8. Use `boxplot()` to plot the weights of the Male individuals.
   Can you transform weight, e.g., `sqrt(Weight) ~ Year`?
   Interpret the results. Do similar boxplots for the t-tests of the
   previous question.

    ```{r brfss-boxplot, fig.width=5, fig.height=5}
    boxplot(Weight ~ Year, brfss, subset = Sex == "Male",
            main="Males")
    ```

9. Use `hist()` to plot a histogram of weights of the 1990 Female individuals.

    ```{r brfss-hist, fig.width=5, fig.height=5}
    hist(brfss_1990[brfss_1990$Sex == "Female", "Weight"],
         main="Females, 1990", xlab="Weight" )
    ```

[BRFSS]: http://www.cdc.gov/brfss/about/index.htm

### *BONUS*: ggplot2


```{r}
library(ggplot2)
```

- http://docs.ggplot2.org

'Grammar of graphics'

- Specify data and 'aesthetics' (`aes()`) to be plotted
- Add layers (`geom_*()`) of information

Clean it by coercing `Year` to factor. A factor is a categorical variable. In this case, our data have only two years represented, so we will treat these two years as "groups" or categories.

```{r}
brfss$Year <- factor(brfss$Year)
```

Let's make a couple of subsets of data to work with. First, let's subset to get only males in 2010.

```{r males2010}
brfss2010Male = subset(brfss,Sex=='Male' & Year=='2010')
```

and make an "only female" subset.

```{r femalesOnly}
brfssFemale = subset(brfss,Sex=='Female')
```


    ```{r, warning=FALSE}
    ggplot(brfss2010Male, aes(x=Height, y=Weight)) +
        geom_point() +
        geom_smooth(method="lm")
    ```
    
- Capture a plot and augment it

    ```{r, warning=FALSE}
    plt <- ggplot(brfss2010Male, aes(x=Height, y=Weight)) +
        geom_point() +
        geom_smooth(method="lm")
    plt + labs(title = "2010 Male")
    ```
    
- Use `facet_*()` for layouts

    ```{r, warning=FALSE}
    ggplot(brfssFemale, aes(x=Height, y=Weight)) +
        geom_point() + geom_smooth(method="lm") +
        facet_grid(. ~ Year)
    ```
        
- Choose display to emphasize relevant aspects of data

    ```{r, warning=FALSE}
    ggplot(brfssFemale, aes(Weight, fill=Year)) +
        geom_density(alpha=.2)
    ```


