---
title: 'Multivariate data analysis: Acute Leukemia'
output: html_document
---

## ALL Phenotypic Data

This data comes from an (old) Acute Lymphoid Leukemia microarray data
set. The actual data are available at [THIS LINK](https://raw.githubusercontent.com/seandavi/ITR/master/ALL-phenoData.csv) or by downloading 
using R.

```{r}
download.file('https://raw.githubusercontent.com/seandavi/ITR/master/ALL-phenoData.csv',
              destfile = 'ALL-phenoData.csv')
```

Choose the file that contains ALL (acute lymphoblastic leukemia)
patient information and input the date using `read.csv()`; for
`read.csv()`, use `row.names=1` to indicate that the first column
contains row names.

```{r ALL-choose, eval=FALSE}
path <- file.choose()    # look for ALL-phenoData.csv
```

<!--
```{r echo=FALSE}
path <- "ALL-phenoData.csv"
```
-->

```{r ALL-input}
stopifnot(file.exists(path))
pdata <- read.csv(path, row.names=1)
```

Check out the help page `?read.delim` for input options. The exercises
use `?read.csv`; Can you guess why? Explore basic properties of the
object you've created, for instance...

```{r ALL-properties}
class(pdata)
colnames(pdata)
dim(pdata)
head(pdata)
summary(pdata$sex)
summary(pdata$cyto.normal)
```

Remind yourselves about various ways to subset and access columns of a
data.frame

```{r ALL-subset}
pdata[1:5, 3:4]
pdata[1:5, ]
head(pdata[, 3:5])
tail(pdata[, 3:5], 3)
head(pdata$age)
head(pdata$sex)
head(pdata[pdata$age > 21,])
```

It seems from below that there are 17 females over 40 in the data
set. However, some individuals have `NA` for the age and / or sex, and
these `NA` values propagate through some computations. Use `table()`
to summarize the number of females over 40, and the number of samples
for which this classification cannot be determined. When _R_
encounters an `NA` value in a subscript index, it introduces an `NA`
into the result. Observe this (rows of `NA` values introduced into the
result) when subsetting using `[` versus using the `subset()`
function.

```{r ALL-subset-NA}
idx <- pdata$sex == "F" & pdata$age > 40
table(idx, useNA="ifany")
dim(pdata[idx,])           # WARNING: 'NA' rows introduced
tail(pdata[idx,])
dim(subset(pdata, idx))    # BETTER: no NA rows
dim(subset(pdata, (sex == "F") & (age > 40)))  # alternative
tail(subset(pdata,idx))

## robust `[`: exclude NA values
dim(pdata[idx & !is.na(idx),])
```

Use the `mol.biol` column to subset the data to contain just
individuals with 'BCR/ABL' or 'NEG', e.g.,

```{r ALL-BCR/ABL-subset}
bcrabl <- subset(pdata, mol.biol %in% c("BCR/ABL", "NEG"))
```

The `mol.biol` column is a factor, and retains all levels even after
subsetting. It is sometimes convenient to retain factor levels, but in
our case we use `droplevels()` to removed unused levels

```{r ALL-BCR/ABL-drop-unused}
bcrabl$mol.biol <- droplevels(bcrabl$mol.biol)
```

The `BT` column is a factor describing B- and T-cell subtypes

```{r ALL-BT}
levels(bcrabl$BT)
```

How might one collapse B1, B2, ... to a single type B, and likewise
for T1, T2, ..., so there are only two subtypes, B and T? One strategy
is to replace two-letter level (e.g., `B1`) with the single-letter
level (e.g., `B`). Do this using `substring()` to select the first
letter of level, and update the previous levels with the new value
using `levels<-`.

```{r ALL-BT-recode}
table(bcrabl$BT)
levels(bcrabl$BT) <- substring(levels(bcrabl$BT), 1, 1)
table(bcrabl$BT)
```

Use `aggregate()` to count the number of samples with B- and T-cell
types in each of the BCR/ABL and NEG groups

```{r ALL-BCR/ABL-BT}
aggregate(rownames(bcrabl) ~ BT + mol.biol, bcrabl, length)
```

Use `aggregate()` to calculate the average age of males and females in
the BCR/ABL and NEG treatment groups.

```{r ALL-aggregate}
aggregate(age ~ mol.biol + sex, bcrabl, mean)
```

Use `t.test()` to compare the age of individuals in the BCR/ABL versus
NEG groups; visualize the results using `boxplot()`. In both cases,
use the `formula` interface. Consult the help page `?t.test` and re-do
the test assuming that variance of ages in the two groups is
identical. What parts of the test output change?

```{r ALL-age}
t.test(age ~ mol.biol, bcrabl)
boxplot(age ~ mol.biol, bcrabl)
```

# Multivariate analysis

This is a classic microarray experiment. Microarrays
consist of 'probesets' that interogate genes for their level of
expression. In the experiment we're looking at, there are 12625
probesets measured on each of the 128 samples. The raw expression
levels estimated by microarray assays require considerable
pre-processing, the data we'll work with has been pre-processed.

## Input and setup

Start by finding the expression data file on disk.

<!--
```{r echo=FALSE}
path <- "ALL-expression.csv"
stopifnot(file.exists(path))
```
-->

```{r eval=FALSE}
path <- file.choose()          # look for ALL-expression.csv
stopifnot(file.exists(path))
```

The data is stored in 'comma-separate value' format, with each
probeset occupying a line, and the expression value for each sample in
that probeset separated by a comma. Input the data using
`read.csv()`. There are three challenges:

1. The row names are present in the first column of the data. Tell _R_
   this by adding the argument `row.names=1` to `read.csv()`.
2. By default, _R_ checks that column names do not look like numbers,
   but our column names _do_ look like numbers. Use the argument
   `check.colnames=FALSE` to over-ride _R_'s default.
3. `read.csv()` returns a `data.frame`. We could use a `data.frame` to
   work with our data, but really it is a `matrix()` -- the columns
   are of the same type and measure the same thing. Use `as.matrix()`
   to coerce the `data.frame` we input to a `matrix`.

```{r }
exprs <- read.csv(path, row.names=1, check.names=FALSE)
exprs <- as.matrix(exprs)
class(exprs)
dim(exprs)
exprs[1:6, 1:10]
range(exprs)
```

We'll make use of the data describing the samples

<!--
```{r echo=FALSE}
path <- "ALL-phenoData.csv"
stopifnot(file.exists(path))
```
-->

```{r ALL-phenoData.csv-clustering-student, eval=FALSE}
path <- file.choose()         # look for ALL-phenoData.csv
stopifnot(file.exists(path))
```

```{r}
pdata <- read.csv(path, row.names=1)
class(pdata)
dim(pdata)
head(pdata)
```

Some of the results below involve plots, and it's convenient to choose
pretty and functional colors. We use the [RColorBrewer][]
package; see [colorbrewer.org][]

[RColorBrewer]: https://cran.r-project.org/?package=RColorBrewer
[colorbrewer.org]: http://colorbrewer.org

```{r}
library(RColorBrewer)  ## not available? install package via RStudio
highlight <- brewer.pal(3, "Set2")[1:2]
```

`highlight' is a vector of length 2, light and dark green.

For more options see `?RColorBrewer` and to view the predefined
palettes `display.brewer.all()`

## Cleaning

We'll add a column to `pdata`, derived from the `BT` column, to
indicate whether the sample is B-cell or T-cell ALL.

```{r ALL-BorT}
pdata$BorT <- factor(substr(pdata$BT, 1, 1))
```

Microarray expression data is usually represented as a matrix of genes
as rows and samples as columns. Statisticians usually think of their
data as samples as rows, features as columns. So we'll transpose the
expression values

```{r}
exprs <- t(exprs)
```

Confirm that the `pdata` rows correspond to the `exprs` rows.

```{r}
stopifnot(identical(rownames(pdata), rownames(exprs)))
```

## Unsupervised machine learning -- multi-dimensional scaling

Reduce high-dimensional data to lower dimension for visualization.

Calculate distance between _samples_ (requires that the expression
matrix be transposed).

```{r}
d <- dist(exprs)
```

Use the `cmdscale()` function to summarize the distance matrix into
two points in two dimensions.

```{r}
cmd <- cmdscale(d)
```

Visualize the result, coloring points by B- or T-cell status

```{r}
plot(cmd, col=highlight[pdata$BorT])
```
